---
title: "The Hitchhiker’s Guide to untargeted lipidomics analysis: Practical guidelines"
author: "D. Smirnov, P. Mazin, M. Osetrova, E. Stekolshchikova, E. Khrameeva"
date: "8/18/2021"
output: 
  github_document:
    pandoc_args: "--webtex"

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
Here we present step-by-step guide to bioinformatic analysis of untargeted LS-MS lipidomic data. This manual reproduces the key steps described in the manuscript:

* Data importing

* Lipid signal quantification (peak peaking, peak alignment, peak grouping and abundance quantification)

* Filtering and normalization

* Visualization


## Package import
Loading the packages required for analysis
```{r, results='hide', message=FALSE, warning=FALSE}
library(xcms)
library(ggplot2)
library(DT)
library(IPO)
library(mixOmics)
```


## Data import
We will demonstrate the key concepts of LC-MS untargeted lipidomic analysis on the example of human fibroblasts dataset (ST001691 study from https://www.metabolomicsworkbench.org/). 
The raw files converted into the .mzXML format can be downloaded in the current directory using the following code:
```{r}
#url <- "http://arcuda.skoltech.ru/~d.smirnov/sampledata.tar.gz"
#download.file(url, destfile = 'sampledata.tar.gz')
#untar('sampledata.tar.gz')
```

For convenience, raw MS files located in the `sampledata/` are organized into the two subfolders according the treatment groups. 
```{r}
mzfiles <- list.files('sampledata/', recursive = TRUE, full.names = TRUE, pattern = '.mzXML')
group <- unlist(lapply(strsplit(mzfiles,"/"), function (x) x[[3]]))
pd <- data.frame(sample_name = sub(basename(mzfiles), pattern = ".mzXML", replacement = "", fixed = TRUE), 
                 sample_group = group, 
                 stringsAsFactors = FALSE)

knitr::kable(pd)
```

Now .mzXML files can be imported via `readMSData` function
```{r}
raw_data <- readMSData(files = mzfiles, 
                       pdata = new("NAnnotatedDataFrame", pd), 
                       mode = "onDisk", 
                       msLevel = 1, 
                       verbose = T, 
                       centroided = T)
```

## Peak picking
Feature detection `CentWave` algorithm based on continuous wavelet transformation allows to distinguish lipid peaks from background noise. To perform peak calling we need to set up CentWave parameters first.
```{r}
cwp <- CentWaveParam(peakwidth = c(9.5, 36),
                     ppm = 11.5,
                     noise = 0, 
                     snthresh = 10, 
                     mzdiff = -0.001, 
                     prefilter = c(3, 100), 
                     mzCenterFun = "wMean", 
                     integrate = 1, 
                     fitgauss = FALSE)
```

Once the parameters are set one can proceed to chromatographic peak detection using `findChromPeaks` function.
```{r, message=FALSE}
xset <- findChromPeaks(raw_data, param = cwp)
```

## Peak alignment
```{r}
arp <- ObiwarpParam(distFun = "cor_opt", 
                    binSize = 1, 
                    response = 1, 
                    gapInit = 0.32, 
                    gapExtend = 2.688, 
                    factorDiag = 2, 
                    factorGap = 1,
                    localAlignment = FALSE)
```

```{r, message=FALSE}
xset <- adjustRtime(xset, param = arp)
```

## Peak grouping
```{r}
pdp <- PeakDensityParam(sampleGroups = xset$sample_group, 
                        bw = 0.879999999999999, 
                        binSize = 0.02412, 
                        minFraction = 0.000001, 
                        minSamples = 1, 
                        maxFeatures = 50)
```

```{r}
xset <- groupChromPeaks(xset, param = pdp)
```

## Selection of parameters for peak picking, alignment, and grouping
In the code sections above we provide parameters optimized for 
```{r}
peakpickingParameters <- getDefaultXcmsSetStartingParams('centWave')
peakpickingParameters$min_peakwidth = c(0,10)
peakpickingParameters$max_peakwidth = c(10,30)
peakpickingParameters$ppm = c(0,10)
```

Please note that optimizing peak calling parameters with `optimizeXcmsSet` function is a quite computationally intensive operation. For simplicity and saving time, in the sections above we used already optimized parameters but ...
```{r}
#resultPeakpicking <- optimizeXcmsSet(files = mzfiles, 
#                                     params = peakpickingParameters, 
#                                     nSlaves = 0, 
#                                     subdir = NULL)
```

```{r}
#optimizedXcmsSetObject <- resultPeakpicking$best_settings$xset
```

```{r}
#retcorGroupParameters <- getDefaultRetGroupStartingParams()
#resultRetcorGroup <- optimizeRetGroup(xset = optimizedXcmsSetObject, 
#                                      params = retcorGroupParameters,
#                                      nSlaves = 0,
#                                      subdir = NULL)

#writeRScript(resultPeakpicking$best_settings$parameters, 
#             resultRetcorGroup$best_settings, 1)

```

## Imputation of missing values
Unfortunately, peak peaking algorithm may produces a sufficient numbers of NaNs for those samples in which it wasn't able to identify MS peaks. We will try to impute missing chromatographic peaks within samples using `fillChromPeaks` function.
```{r}
xset <- fillChromPeaks(xset)
```

## Data export

```{r, message=FALSE}
pks <- chromPeaks(xset)
grs <- featureDefinitions(xset)
mtx <- featureValues(xset, method="maxint", value="into", filled=T) 

knitr::kable(head(mtx))
```

## Filtering of peaks
The code below removes all peaks that possess more than 30% of NaNs across samples.
```{r, message=FALSE}
th <- 0.3
peaks.nas <- apply(mtx, 1, function (x) sum(is.na(x)))
mtx <- mtx[(peaks.nas/ncol(mtx)) < th, ]
```

## Normalization

## Annotation

We will load the 
```{r}
source("src/rt-mz.annotator.R")
```

```{r}
grs <- as.data.frame(grs)
grs.short <- dplyr::select(grs, mzmed, rtmed)
grs.short$id <- paste(round(grs.short$mzmed, 3), round(grs.short$rtmed, 3), sep = '_')
grs.short <- grs.short[,c(3,1,2)]
colnames(grs.short) <- c('id', 'mz', 'rt')
knitr::kable(head(grs.short))
```

```{r, message=FALSE, echo=FALSE}
ann = annotateByMass(grs.short, db = LMDB)
```

```{r}
knitr::kable(head(ann))
```

## Downstream analysis
 
### Principal Component Analysis (PCA)
PCA projects original matrix of lipid abundances into low dimensional space. For a given matrix $X_{n \times m}$, where samples are rows and samples are columns, principal component vectors can be defined by finding eigenvectors of the following sample covariance matrix $S$:

![\begin{align*}
S = \frac{1}{n-1} X^{T} C_n X
\end{align*}
](https://render.githubusercontent.com/render/math?math=%5Cdisplaystyle+%5Cbegin%7Balign%2A%7D%0AS+%3D+%5Cfrac%7B1%7D%7Bn-1%7D+X%5E%7BT%7D+C_n+X%0A%5Cend%7Balign%2A%7D%0A)

where $C_n = I_n - \frac{1}{n}1_{n} 1^{T}_{n}$ is a centering matrix, $I_n$ represents an identity matrix of size $n$. 

To calculate principal components we will use base R function `prcomp`
```{r, message=FALSE}
pca <- prcomp(t(na.omit(mtx)), center = TRUE, scale. = TRUE)
```

To visualize relationships between samples in a new low dimensional space we will plot PC1 and PC2 against each other.

```{r, message=FALSE}
Y <- pd$sample_group 
pca.data <- data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 2], class = Y)
ggplot(data = pca.data, aes_string(x = "PC1", y = "PC2", color = "class", shape = "class")) + 
    geom_point(size = 4) +
    theme_light()
```

### Partial Least-Squares Discriminant Analysis (PLS-DA)

Mathematically, PLS-DA principal components can be obtained in a similar to PCA manner, as eigenvectors of a matrix of covariances between $X$ and $Y$: 

![\begin{align*}
S = S_{xy} S_{yx} = \frac{1}{n-1} X^{T} C_n Y \times \frac{1}{n-1} Y^{T} C_n X = \frac{1}{(n-1)^2} X^{T} C_n Y Y^{T} C_n X
\end{align*}
](https://render.githubusercontent.com/render/math?math=%5Cdisplaystyle+%5Cbegin%7Balign%2A%7D%0AS+%3D+S_%7Bxy%7D+S_%7Byx%7D+%3D+%5Cfrac%7B1%7D%7Bn-1%7D+X%5E%7BT%7D+C_n+Y+%5Ctimes+%5Cfrac%7B1%7D%7Bn-1%7D+Y%5E%7BT%7D+C_n+X+%3D+%5Cfrac%7B1%7D%7B%28n-1%29%5E2%7D+X%5E%7BT%7D+C_n+Y+Y%5E%7BT%7D+C_n+X%0A%5Cend%7Balign%2A%7D%0A)

where $C_n$ and $n$ represent a centering matrix and a total number of samples, respectively.

```{r, message=FALSE}
X <- t(na.omit(mtx))
Y <- as.factor(Y)
plsda.model <- plsda(X, Y, ncomp = 2)
plotIndiv(plsda.model, ind.names = FALSE, legend=TRUE, ellipse = TRUE)
```

```{r, message=FALSE}
plsda.contributions <- selectVar(plsda.model, comp = 1)$value
```

## References

## Software used
```{r}
sessionInfo()
```









