---
title: "The Hitchhiker’s Guide to untargeted lipidomics analysis: Practical guidelines"
author: "D. Smirnov, P. Mazin, M. Osetrova, E. Stekolshchikova, E. Khrameeva"
date: "8/18/2021"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction


## Package import
Loading the packages required for analysis
```{r, results='hide', message=FALSE, warning=FALSE}
library(xcms)
library(ggplot2)
library(DT)
library(IPO)
library(mixOmics)
```


## Data import
We will demonstrate the key concepts of LC-MS untargeted lipidomic analysis on the example of human fibroblasts dataset (ST001691 study from https://www.metabolomicsworkbench.org/). 
The raw files converted into the .mzXML format can be downloaded in the current directory using the following code:
```{r}
#url <- "http://arcuda.skoltech.ru/~d.smirnov/sampledata.tar.gz"
#download.file(url, destfile = 'sampledata.tar.gz')
#untar('sampledata.tar.gz')
```

For convenience, raw MS files located in the `sampledata/` are organized into the two subfolders according the treatment groups. 
```{r}
mzfiles <- list.files('sampledata/', recursive = TRUE, full.names = TRUE, pattern = '.mzXML')
group <- unlist(lapply(strsplit(mzfiles,"/"), function (x) x[[3]]))
pd <- data.frame(sample_name = sub(basename(mzfiles), pattern = ".mzXML", replacement = "", fixed = TRUE), 
                 sample_group = group, 
                 stringsAsFactors = FALSE)

knitr::kable(pd)
```

Now .mzXML files can be imported via `readMSData` function
```{r}
raw_data <- readMSData(files = mzfiles, 
                       pdata = new("NAnnotatedDataFrame", pd), 
                       mode = "onDisk", 
                       msLevel = 1, 
                       verbose = T, 
                       centroided = T)
```

## Peak picking
Feature detection `CentWave` algorithm based on continuous wavelet transformation allows to distinguish lipid peaks from background noise. To perform peak calling we need to set up CentWave parameters first.
```{r}
cwp <- CentWaveParam(peakwidth = c(9.5, 36),
                     ppm = 11.5,
                     noise = 0, 
                     snthresh = 10, 
                     mzdiff = -0.001, 
                     prefilter = c(3, 100), 
                     mzCenterFun = "wMean", 
                     integrate = 1, 
                     fitgauss = FALSE)
```

Once the parameters are set one can proceed to chromatographic peak detection using `findChromPeaks` function.
```{r, message=FALSE}
xset <- findChromPeaks(raw_data, param = cwp)
```

## Peak alignment
```{r}
arp <- ObiwarpParam(distFun = "cor_opt", 
                    binSize = 1, 
                    response = 1, 
                    gapInit = 0.32, 
                    gapExtend = 2.688, 
                    factorDiag = 2, 
                    factorGap = 1,
                    localAlignment = FALSE)
```

```{r, message=FALSE}
xset <- adjustRtime(xset, param = arp)
```

## Peak grouping
```{r}
pdp <- PeakDensityParam(sampleGroups = xset$sample_group, 
                        bw = 0.879999999999999, 
                        binSize = 0.02412, 
                        minFraction = 0.000001, 
                        minSamples = 1, 
                        maxFeatures = 50)
```

```{r}
xset <- groupChromPeaks(xset, param = pdp)
```

## Selection of parameters for peak picking, alignment, and grouping
```{r}
peakpickingParameters <- getDefaultXcmsSetStartingParams('centWave')
peakpickingParameters$min_peakwidth = c(0,10)
peakpickingParameters$max_peakwidth = c(10,30)
peakpickingParameters$ppm = c(0,10)
```

```{r}
#resultPeakpicking <- optimizeXcmsSet(files = mzfiles, 
#                                     params = peakpickingParameters, 
#                                     nSlaves = 0, 
#                                     subdir = NULL)
```


## Imputation of missing values
```{r}
xset <- fillChromPeaks(xset)
```

## Data export

```{r, message=FALSE}
pks <- chromPeaks(xset)
grs <- featureDefinitions(xset)
mtx <- featureValues(xset, method="maxint", value="into", filled=T) 

knitr::kable(head(mtx))
```

## Filtering of peaks
Removing peaks with NaN
```{r, message=FALSE}
th <- 0.3
```

```{r, message=FALSE}
peaks.nas <- apply(mtx, 1, function (x) sum(is.na(x)))
mtx <- mtx[(peaks.nas/ncol(mtx)) < th, ]
```

## Normalization



## Downstream analysis

## Principal Component Analysis (PCA)

```{r, message=FALSE}
pca <- prcomp(t(na.omit(mtx)), center = TRUE, scale. = TRUE)
```

```{r, message=FALSE}
Y <- pd$sample_group 
```

```{r, message=FALSE}
pca.data <- data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 2], class = Y)
ggplot(data = pca.data, aes_string(x = "PC1", y = "PC2", color = "class")) + geom_point()
```

## Partial Least-Squares Discriminant Analysis (PLS-DA)
```{r, message=FALSE}
X <- t(na.omit(mtx))
Y <- as.factor(Y)
plsda.model <- plsda(X, Y, ncomp = 2)
plotIndiv(plsda.model, ind.names = FALSE, legend=TRUE, ellipse = TRUE)
```

```{r, message=FALSE}
plsda.contributions <- selectVar(plsda.model, comp = 1)$value
```

## Software used
```{r}
sessionInfo()
```









